---
output: github_document
---

<!-- README.md is generated from README.Rmd -->

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## visProteomics

**visProteomics** helps with the exploration and visualization of proteomics data obtained with the process of fractionation. Fractionation of samples is often used in proteomics to improve signal-to-noise ratio and proteome coverage. **visProteomics** explores attributes of proteins detected in different fractions through both graphical and tabular representation.

Venn/Euler diagrams are widely used for data summarization and there are many plotting tools available. **visProteomics** defines a complementary tabular function. The resulting data frame summarizes attributes of elements found in different regions of the Venn diagram. This generic function can be used for any data set.


## Installation

```{r install_github, eval = FALSE}
devtools::install_github("ivabudimir/visProteomics")
```

For installation of vignettes together with the package, add "build_vignettes=TRUE".

## Usage

Please check [visProteomics](https://ivabudimir.github.io/visProteomics/) website.

Alternatively, read vignettes *merge-fractions* and *vis-fractions* for the exploration and visualization of a fractionated proteomics sample. For the tabular Venn function, read *explore-venn* vignette.

## License

visProteomics is open source software, licensed under 
[GPL-3](https://github.com/ivabudimir/visProteomics/blob/master/LICENSE).
